% The Attractor Atlas - methods paper.
% Venue-agnostic skeleton: plain article + natbib. Switch to ACM sigconf
% (GECCO) or the AIP template (Chaos) at submission time.
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[colorlinks=true,linkcolor=blue!50!black,citecolor=blue!50!black,urlcolor=blue!50!black]{hyperref}

\graphicspath{{../gallery/}}

\newcommand{\dky}{D_{\mathrm{KY}}}
\newcommand{\conj}[1]{\bar{#1}}

\title{Quality-Diversity Search with Catalog-Grounded Novelty\\
Discovers Families of Equivariant Strange Attractors}
\author{Sakeeb Rahman\thanks{Code, data, and an interactive viewer:
\url{https://github.com/Sakeeb91/attractor-atlas},
\url{https://sakeeb91.github.io/attractor-atlas/}.}}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
Random search has been generating chaotic systems since the early 1990s, and
evolutionary methods now breed them by the tens of thousands; what neither
provides is a reason to look at any particular specimen. We describe a
search whose objective is \emph{measured difference from everything already
known}: a quality-diversity (MAP-Elites) loop over a $C_n$-equivariant
family of three-dimensional polynomial flows, in which each candidate orbit
is reduced to a geometry-only shape fingerprint and scored by its distance
to the 135-system \textsf{dysts} catalog of published chaotic attractors.
Cells of the archive are indexed by symmetry order, Lyapunov exponent, and
body aspect, so the search illuminates the family rather than converging.
A first run evaluated 4{,}608 candidate systems in 127 seconds and retained
62 elites. We certify three exhibits --- Naiad ($C_2$), Aurelia ($C_3$),
and Cassiopea ($C_4$) --- with full Lyapunov spectra and error bars
($\lambda_1 = 0.294\pm0.009$, $0.230\pm0.009$, $0.523\pm0.008$), fractal
dimensions, equilibrium analyses, bifurcation scans, and a homoclinic hunt
that finds near-Shilnikov connections (return distances down to
$7.5\times10^{-5}$) threading two of the three. We validate the fingerprint
metric itself with positive and negative controls and an explicit noise
floor; all three systems sit 3--5 noise floors from anything in the
catalog. The exhibits are evidence for the method: catalog-grounded novelty
turns ``yet another chaotic system'' into a measurable, falsifiable claim
of isolation in shape space.
\end{abstract}

\section{Introduction}\label{sec:intro}

Strange attractors have been discovered mostly one at a time. Lorenz found
his by accident in a truncated convection model \citep{lorenz1963}; for
three decades afterwards each new chaotic flow was an event, named for its
finder. Sprott industrialized the process in 1994 by integrating millions
of random quadratic systems and keeping the nineteen algebraically simplest
chaotic ones \citep{sprott1994}, having already automated the analogous
hunt over two-dimensional maps with an explicitly aesthetic filter
\citep{sprott1993}. The genre this spawned eventually saturated: so many
cosmetic variants of known systems were published that Sprott himself
proposed editorial standards --- a new chaotic system should be simpler
than anything known with the same behavior, or show behavior of genuine
scientific interest \citep{sprott2011}. Modern evolutionary methods have
only widened the firehose: genetic programming evolves hundreds of
Lorenz-like variants by maximizing the largest Lyapunov exponent
\citep{panDas2015}, and the Panda corpus bred $2\times10^4$ chaotic
ordinary differential equations from catalog seeds as training data for a
forecasting foundation model \citep{lai2025panda}. Generating chaotic
systems is cheap. What remains expensive is a \emph{reason to look at any
particular one}.

This paper proposes that the missing objective is measured novelty. The
machine-readable \textsf{dysts} catalog \citep{gilpin2021} collects 135
published chaotic systems --- a century of the bestiary in one place ---
and makes a new kind of search target possible: not ``is this chaotic?''
(easy, and no longer interesting on its own) but ``how far is this, as a
geometric object, from everything anyone has published?'' We reduce every
candidate attractor to a cheap shape fingerprint (a D2 pairwise-distance
histogram \citep{osada2002} plus gross-anatomy features), define novelty
as the minimum fingerprint distance to the catalog, and make that number
the selection signal.

The second ingredient is a search algorithm that does not converge.
Optimizing novelty alone collapses easily onto degenerate weirdness ---
our own first run surfaced a nearly flat, barely chaotic sheet as its most
novel specimen --- and optimizing chaos strength alone reproduces the
firehose. Quality-diversity methods \citep{lehmanStanley2011,pugh2016},
specifically MAP-Elites \citep{mouretClune2015}, hold a map of cells and
keep the best specimen per cell; we index cells by symmetry order,
Lyapunov exponent, and body aspect, so the output is an \emph{atlas} of
distinct, certified, maximally unfamiliar attractors rather than one
optimum.

The third ingredient is structure. Rather than searching free-form
equations, we search a single $C_n$-equivariant family of
three-dimensional polynomial flows (Section~\ref{sec:family}), built from
the unique resonant coupling $\gamma\conj{w}^{\,n-1}$ of equivariant
bifurcation theory \citep{golubitskyStewartSchaeffer1988} joined to cubic
axial dynamics. The family is parameterized by the rotation order $n$, so
two-, three-, four-, and five-fold symmetric attractors live in one
seven-dimensional parameter box, and every find inherits an interpretable
anatomy: a single on-axis saddle-focus whose unstable spiral plane and
strongly stable axis organize a symmetric recirculation.

We present three certified exhibits --- Naiad ($C_2$), Aurelia ($C_3$),
Cassiopea ($C_4$) --- but we emphasize that, per Sprott's standard, the
exhibits are not the claim. The claim is the method: to our knowledge this
is the first quality-diversity search over a family of dynamical systems,
and the first use of catalog-grounded novelty as the selection objective
for discovering chaotic flows. Our contributions:
\begin{enumerate}
  \item a novelty objective grounded in a scientific catalog, with the
        metric itself validated by positive and negative controls and an
        explicit noise floor (Section~\ref{sec:map});
  \item an equivariant polynomial ansatz that turns symmetry order into a
        searchable discrete dimension (Section~\ref{sec:family});
  \item a MAP-Elites archive over symmetry $\times$ dynamics descriptors
        with a batched evaluator (4{,}608 candidates in 127\,s on a
        laptop; Section~\ref{sec:run});
  \item a certification standard for machine-discovered systems: full
        spectra with error bars over integrator settings and initial
        conditions, equilibrium analysis, bifurcation scans, and a
        homoclinic hunt (Sections~\ref{sec:convergence}--\ref{sec:shilnikov});
  \item three certified, named exhibits and two measured surprises ---
        chaos strength is not monotonic in symmetry order, and
        near-homoclinic Shilnikov windows thread two of the three
        families (Section~\ref{sec:nonmonotonic}).
\end{enumerate}

\section{Related work}\label{sec:related}

\paragraph{Symmetric chaos.}
Field and Golubitsky built chaotic planar \emph{maps} equivariant under
cyclic, dihedral, and crystallographic groups --- the symmetric icons and
quilts --- explicitly in pursuit of visual beauty
\citep{fieldGolubitsky2009}; our family is a flow-analogue of that program.
The algebra we use is theirs in the broader sense: for a planar system with
$\mathbb{Z}_q$ rotation symmetry, the lowest-order non-axisymmetric
equivariant term is the conjugate power $\conj{w}^{\,q-1}$
\citep{golubitskyStewartSchaeffer1988}; we do not extend that theory, we
deploy it globally as a polynomial ansatz. The admissible symmetry groups
of attractors, and the distinction between instantaneous and time-averaged
symmetry, were classified by \citet{melbourneDellnitzGolubitsky1993} and
\citet{ashwinMelbourne1994}; our attractors carry the full $C_n$ symmetry
on average.

\paragraph{$C_n$-equivariant 3-D flows exist already.}
Discrete rotational symmetry in three-dimensional chaos is not new, and
our novelty claims are calibrated against that literature.
\citet{mirandaStone1993} quotiented the Lorenz system to the
``proto-Lorenz'' and lifted it to $n$-fold covers with $n$-eared
attractors; \citet{letellierGilmore2001} and the monograph of
\citet{gilmoreLetellier2007} developed the systematic theory of such
cover/image systems. Thomas's cyclically symmetric system carries a
three-fold rotation about the space diagonal \citep{thomas1999}, and the
multi-scroll literature engineers rotationally arranged attractors at will
\citep{luChen2006}, including modern rotation-symmetric multi-wing
constructions around stable equilibria \citep{yang2021multiwing}. What
separates our family from all of these: cover systems re-present the
dynamics of a known image (away from the rotation axis a cover is locally
diffeomorphic to its image, and the lifted vector field is generically
singular on the axis), and multi-scroll systems are designed top-down with
non-smooth nonlinearities. Our systems are single, globally polynomial,
smooth vector fields, equivariant by construction rather than by lifting,
not covers of anything catalogued, with the symmetry axis passing through
the attractor's organizing saddle-focus --- and they were found by search,
not built.

\paragraph{The new-chaotic-system genre.}
\citet{sprott1994} is the ancestor of automated chaotic-flow discovery;
\citet{sprott2011} is the editorial corrective to the flood it unleashed,
and the reason this is a methods paper. One-equilibrium chaotic flows are
an established genre --- the minimal jerk flow \citep{sprott1997}, the
one-stable-equilibrium system of \citet{wangChen2012}, and twenty-three
relatives \citep{molaie2013} --- so a single equilibrium is not by itself
remarkable; what matters here is the geometry, a saddle-focus whose 2-D
unstable spiral and 1-D strongly stable axis organize the entire symmetric
recirculation, the configuration of Shilnikov's theorem
\citep{shilnikov1965}.

\paragraph{Automated discovery of chaotic systems.}
The lineage runs from Sprott's automatic aesthetic generation of map
attractors \citep{sprott1993}, through genetic-algorithm acceleration
\citep{goertzel1995}, to genetic programming over Lorenz-like vector
fields maximizing $\lambda_1$ \citep{panDas2015}, and most recently the
evolutionary generation of $2\times10^4$ chaotic ODEs as training data
\citep{lai2025panda}. None of these searches for novelty against the
published record; selection is chaoticity or chaos strength, and the
output is a pile rather than a map. The nearest methodological neighbor
outside dynamics is intrinsically-motivated discovery of diverse patterns
in cellular automata \citep{reinke2020}.

\paragraph{Catalogs and fingerprints.}
\textsf{dysts} \citep{gilpin2021} supplies the reference set (135
continuous systems in the version we fingerprint) with standardized
integration; Gilpin's own feature-space embedding of the catalog is the
closest precedent for our shape-space map, with the difference that our
features are geometry-only point-cloud descriptors --- the D2 shape
distribution of \citet{osada2002} plus PCA eigenvalue ratios and a fill
factor --- rather than forecasting-oriented statistics.

\paragraph{Quality-diversity.}
Novelty search \citep{lehmanStanley2011}, MAP-Elites
\citep{mouretClune2015}, and the QD framing \citep{pugh2016} come from
evolutionary robotics. The domain transfer here is twofold: novelty is
measured against a \emph{scientific catalog} rather than against the
evolving population, and the archive's cells are indexed by a symmetry
order, which makes the discrete group $C_n$ a searchable behavioral
dimension. We found no prior application of QD or novelty search to the
discovery of dynamical systems.

\section{Method}\label{sec:method}

\subsection{The equivariant family}\label{sec:family}

Write the horizontal coordinates of $\mathbb{R}^3$ as one complex number
$w = x + iy$, so that a rotation of the plane by $2\pi/n$ is multiplication
by $e^{2\pi i/n}$. The search family is
\begin{equation}\label{eq:family}
  \dot w = \bigl[\alpha(z-\beta) + i\omega\bigr]\,w
           \;+\; \gamma\,\conj{w}^{\,n-1},
  \qquad
  \dot z = \mu + \nu z - z^{3} - \lambda\,|w|^{2},
\end{equation}
with seven real parameters
$(\alpha,\beta,\omega,\gamma,\mu,\nu,\lambda)$ and a discrete rotation
order $n \ge 2$. Under $w \mapsto e^{2\pi i/n} w$ the conjugate power
transforms as $\conj{w}^{\,n-1} \mapsto e^{-2\pi i (n-1)/n}\conj{w}^{\,n-1}
= e^{2\pi i/n}\conj{w}^{\,n-1}$, so every term of \eqref{eq:family} is
equivariant: rotating a solution by $2\pi/n$ about the $z$-axis yields
another solution. The coupling $\conj{w}^{\,n-1}$ is the lowest-order term
with this property --- the standard $\mathbb{Z}_n$-resonant monomial of
equivariant bifurcation theory \citep{golubitskyStewartSchaeffer1988} ---
and it is the \emph{only} symmetry-breaking term in the family; everything
else is axisymmetric.

Read \eqref{eq:family} as a machine with three interlocking parts. The
term $i\omega w$ spins the horizontal plane like a turntable. The term
$\alpha(z-\beta)w$ makes the spin unstable or stable depending on
altitude: above $z=\beta$ trajectories spiral outward, below it they are
drawn back in. The vertical equation is a thermostat --- the cubic $-z^3$
caps the altitude while $-\lambda|w|^2$ pushes $z$ down whenever
trajectories wander far from the axis --- and the resonant term warps the
circular spiral into an $n$-lobed one. On the axis $w=0$ the planar
equations vanish and the equilibria are the real roots of
$\mu + \nu z - z^3$; at every parameter set studied below there is exactly
one, a saddle-focus whose unstable eigenplane is exactly the plane
$z = z^\ast$ (the Jacobian is block diagonal there) and whose stable
direction is the axis. Energy pumps in along the axis, spills outward, is
reined back: the recirculating anatomy of Shilnikov-type chaos
\citep{shilnikov1965}, with the symmetry painted on it.

The search box is $\alpha\in[0.5,2.5]$, $\beta\in[-1,1.5]$,
$\omega\in[1,4]$, $\gamma\in[0.3,2]$, $\mu\in[-1,2]$, $\nu\in[-1,2]$,
$\lambda\in[0.5,3]$, with $n \in \{2,3,4,5\}$.

\subsection{Batched evaluation and gates}\label{sec:eval}

Candidates are evaluated in batches of 96 by a vectorized complex RK4
integrator (step $dt=0.01$): an 8{,}000-step transient, then a
25{,}000-step two-orbit Benettin estimate of the largest Lyapunov
exponent (separation $10^{-8}$, renormalized every step), then a
12{,}000-step orbit sample decimated to 1{,}500 points for
fingerprinting. Orbits are culled as dead the moment any coordinate
exceeds $50$ or goes non-finite; culling is checked every 200 steps so a
diverging member of the batch cannot poison the rest. A candidate must
survive the full evaluation with a largest exponent inside the archive's
descriptor range (Section~\ref{sec:mapelites}) to be considered at all.
The evaluator reproduces the independently computed Lyapunov exponent of
the $C_3$ exhibit as a sanity check.

\subsection{Shape fingerprints and catalog-grounded novelty}\label{sec:fingerprint}

A fingerprint must be cheap and invariant to what does not matter:
translation, scale, and rigid rotation. Each orbit sample is centered and
normalized to unit RMS radius, then reduced to (i) the D2 shape
distribution \citep{osada2002} --- the histogram, over 32 bins on
$[0,3]$, of pairwise distances between 400 random orbit points, embedded
by the Hellinger square root --- (ii) the PCA eigenvalue ratios
$e_2/e_1$ and $e_3/e_1$ (how planar, how volumetric), and (iii) the
occupied fraction of a $12^3$ grid over the bounding box (space-filling
versus wiry). The distance between fingerprints is the Euclidean distance
on the histogram block plus one half the Euclidean distance on the
feature block. The novelty of a candidate is its minimum distance to a
reference set containing the 135 fingerprinted systems of the
\textsf{dysts} catalog \citep{gilpin2021} (higher-dimensional systems
projected to their first three coordinates) \emph{plus every elite
already in the archive}, so the search cannot rediscover itself.
Section~\ref{sec:map} validates the metric and measures its noise floor.

\subsection{The MAP-Elites archive}\label{sec:mapelites}

The archive's cells are triples (rotation order $n$; largest Lyapunov
exponent, 7 bins with edges $0.05,0.10,0.15,0.20,0.27,0.35,0.45,0.60$;
vertical aspect ratio of the orbit cloud, 5 bins with edges
$0.2,0.4,0.6,0.85,1.2,2.0$) --- 140 cells in all. Within a cell,
candidates compete on novelty alone: a newcomer replaces the incumbent
only if its catalog-grounded novelty is higher, in the spirit of novelty
search with local competition \citep{lehmanStanley2011,mouretClune2015}.
Each generation draws half its batch as Gaussian mutations of random
elites ($\sigma$ = 8\% of each parameter range, clipped to the box) and
half as fresh uniform samples; the archive persists on disk across runs,
so the search compounds.

\section{Results}\label{sec:results}

\subsection{The atlas run}\label{sec:run}
[Placeholder: 4{,}608 candidates in 127 s, 62 elites.]

\subsection{Three certified systems}\label{sec:systems}
[Placeholder: table of parameters, spectra, $\dky$, equilibria, novelty
distances; the 83\% isolation statistic.]

\subsection{Convergence and error bars}\label{sec:convergence}
[Placeholder: the dt x length x IC grid; Figure convergence.png.]

\subsection{The shape-space map}\label{sec:map}
[Placeholder: fingerprint validation (controls, noise floor) and the MDS
map; Figure fingerprint\_map.png.]

\subsection{Bifurcations and Shilnikov evidence}\label{sec:shilnikov}
[Placeholder: bifurcation scan of Aurelia; homoclinic hunt results;
Figure homoclinic.png.]

\subsection{Chaos strength is not monotonic in symmetry order}\label{sec:nonmonotonic}
[Placeholder: $\lambda_1 = 0.294$ ($C_2$), $0.230$ ($C_3$), $0.523$ ($C_4$)
as an open question.]

\section{Limitations}\label{sec:limitations}
[Placeholder: numerical (not rigorous) certification; dysts is 135 systems,
not the literature; fingerprint crudeness and noise floor; aesthetics
subjectivity.]

\section{Reproducibility}\label{sec:repro}
[Placeholder: one script per claim; live viewer; archive persistence.]

\nocite{*}  % skeleton stage: render the whole verified bibliography
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
